---
title: "Econ/Demog C175 Lab for Week 8: Micro-economic Models of
Fertility and an Investigation of Historical Switzerland
few children?"
output: html_notebook
---


In this lab, we first go further into the cost-of-time model of
fertility, answering questions based on theory and experimentation
with the cost_of_time "app". Then, we look at fertility and
socio-economic data from historical cross-section provinces in
French-speaking Switzerland, trying to see how our models may or may
not be consistent with historical experience.

> Note: The first part of the lab includes some mathematical
> derivations that we did not fully cover in class. You should
> understand what analysis is done (and why), but you don't need to be
> able to derive this yourself. To display the equations you may need
> to click on them or below them in the R Notebook.

> Further note: The graded questions are interspersed throughout the
> lab. They should be numbered consecutively.

> Final note: The computing with data part of the lab is short this
> time and should take less time than the first parts on the
> cost_of_time.


# Part 0. Preliminaries

We begin by downloading the functions to be used in the lab. The data
is actually a built-in R dataset so we don't need to download it.


```{r}
tot = 0
source("http://courses.demog.berkeley.edu/goldstein175/Basket/check_fertility.R")
source("http://courses.demog.berkeley.edu/goldstein175/Basket/check_functions.R")
```

# Part 1. Implications of the Cost-of-Time Model of Fertility Choice



The Cost_of_time app computes the optimal bundle (N, X) of N children
and X-dollars worth of consumer goods. The app plots two choices, one
based on the initial values of $(w, p, c, and I)$ and one for separate
values $(w', p', c', and I'). The app is available at
(http://shiny.demog.berkeley.edu/josh/) under cost_of_time. We
recommend resizing the window of the app and playing with the font
size (cmd-shift+, cmd-shift- on a mac) until the graph is fairly
square.

You can see the effect of changing any single value by changing any
one of the "Parameters to perturb."

You can see how the effect of these changes (e.g., increasing wages)
depends on the values of the other parameters by first changing the
initial parameter values, and then changing one of the "Parameters to
perturb".


## Background to the Model

We imagine potential parents choosing between the number of children
they want to have and the amount of "other goods" they want to
consume.

### Prices

For simplicity, we let there just be one kind of "other goods", which
we denote $X$. And we let the price be $1 per unit of $X$.


The price of children has two components:

1. A fixed dollar expenditure per child $p$ for food, clothing, shelter,
schooling, toys, etc.

2. A time cost of $c$ hours per child. This dollar value of the time
cost depends on the parents wage.

### Income and the Budget Constraint

Parents face a budget constraint. Their household income consists of
'outside income' $I$ and wage income $w L = w (T - c N)$.

1. 'Outside income' $I$ includes sources like interest on assets,
   inheritance, as well as the wage-earnings of someone who cannot
   spend their time with children.

2. Wage income is the hourly wage rate $w$ times the number of hours
   worked $L$. The number of hours worked is equal to the total time
   available for work $T$, minus the hours spent taking care of
   children. (In our simple model, there is no leisure.)

If the division of labor is such that men never take care of children,
and women divide their time between child-care and market work, then
one can think of $I$ as "male earnings" and $w (T- c N)$ as "female
earnings."

The family's expenditures must satisfy the budget constraint
$$
\mbox{Income} = \mbox{Consumption}.
$$
In our model, this means
$$
I + w (T- c N) = X + p N.
$$
We can re-arrange as
$$
I + w T = X + (p + c w) N,
$$
where the left-hand side is now "Full Income" and the right hand side
gives the consumption including the time-cost of children. The full
cost per child is thus $p + c w$.

### Utility

The choice of how many children and how much "other goods" is made so
as to make the parents as happy as possible. We can use the same
Cobb-Douglas formulation that we used for production functinos earlier
in the semester as a convenient model for utility. We let the utility
of a particular choice of $X$ and $N$ be
$$
U(X,N) = X^a N^{1-a},
$$
where $0 < a < 1$.

### The Optimal 'Bundle' of Goods

Parents will choose $X$ and $N$ to maximize $U$, subject to their
budget constraint. There are many methods for solving this constrained
optimization problem, including:

* Numerical search. (We try  all combinations of X and N that satisfy
  the budget constraint). This is what we did in class.

* Substitution. We rewrite our budget constraint so that we express
  one variable in terms of another. Then we substitute this expression
  into the utility function and maximize utility of this single
  variable function. (We will do this below.)

* Lagrange multipliers. This is a general method for constrained
  optimization. Good to know, but beyond the mathematics we're using
  in this course.

### Solving for the Optimum by Substitution

The purpose of showing the derivation is to give you a sense of how
this is done. You don't need to be able to do this yourselves.

1. Express $X$ in terms of $N$.

From the budget constraint,
$$
I + w T = X + (p + c w) N,
$$
we can write
$$
X = (I + w T) - (p + cw) N
$$

2. Substitute into $U$

$$
U(X,N) = X^a N^{1-a},
$$
$$
U(X,N) = U(N) =  \left[ (I + wT) - (p + cw) N \right]^a N^{1-a}
$$

3. Maximization (the rate of change at the optimum should be zero)

At the maximum, $U'(N) = 0$. But it is a bit easier to work with
$\log U$, which will reach a maximimum at the same value of $N$.

$$
\log U(N) = a\log{ \left[(I + wT) - (p + cw) N \right]} + (1-a)\log N
$$

$$
{d\log U(N) \over d N} = {-a (p + cw) \over
\left[(I + wT) - (p + cw) N \right]} + {(1-a) \over N}
$$

When ${d\log U(N) \over d N} = 0$,
$$
{(1-a) \over N} = {a (p + cw) \over
\left[(I + wT) - (p + cw) N \right]}
$$

4. Solving for the optimal values

Solving for $N$ gives us
$$
N^* = (1 - a) {I + w T \over p + cw}.
$$

We obtain $X^*$, by substituting $N^*$ into our earlier expression.
$$
X^* = (I + w T) - (p + cw) N^* = (I + w T) - (1-a)(I + w T) = a (I + w T).
$$
(Because of the properties of the Cobb-Douglas utility function, the
amount of $X$ chosen depends only on full income and is independent of
the price of children.)

### Analysis of the solution

With our solution
$$
N^* = (1 - a) {I + w T \over p + cw}
$$
in hand, we can now ask what effect changes in any of the parameters
will have on the choice of $N$.

For example, increasing $a$ will clearly reduce $N^*$. We can write
$$
{\partial N \over \partial a} < 0
$$

(The symbol $\partial$ indicates a partial derivative, differentiating
only with respect one, individual variable.)

Likewise, we can see that
$$
{\partial N \over \partial I} > 0
$$

And since $p$ and $c$ are both in the denominator,
$$
{\partial N \over \partial c} < 0
$$
and
$$
{\partial N \over \partial p} < 0
$$.

The only effect that is ambiguous in sign is
$$
{\partial N \over \partial w} \lessgtr 0
$$.

Some additional mathematical manipulation shows that

$$
{\partial N \over \partial w} <  0, \mbox{ when } {T\over c} < {I \over p}.
$$.

So, while in class, we questions the usefulness of the cost-of-time
model in making predictions about the effect of a wage increase on
fertility, we see here that the model makes clear predictions about
the other variables. Increases in outside income $I$ unambigously
increase fertility. Increases in the price of children through $c$ or
$p$ unambigously reduce fertility. And changes in taste, by changing
the relative preferences $a$ for children vs. goods, are also
unambiguous in their effect.

For the purposes of this class, you do not need to be able to follow
every step in the mathematical derivation. What is important to
understand is that it is possible to analyze this model and discover
in a quite general way what predictions the model makes and what the
limits of the model are. The tractability of the model is what makes
it so useful for enhancing our understanding.


## Graded Questions on the Cost-of-Time Model of Fertility Choice

For the following questions, we will leave the initial values of $w$, $p$,
$c$, and $I$ unchanged. (Tip: you can always refresh your browser to return
the app to its starting state.)


> 1. Effect of a wage increase, changing $w'$ from $15 to $30:

> If we increase the hourly wage $w'$ from $15 to $30, what happens to
> fertility? [1 sentence with actual values.]

> What happens to utility? [1 sentence with actual values.]

> What happens to consumption of goods? [1 sentence with actual values.]

> 2. If we increase outside income $I'$ from 1000 per week to $1,500 a
>    week, what happens to fertility? [1 sentence with actual values.]

> 3. If we reduce fixed costs $p'$ from $100 per week to $50 per week, what
>    happens to fertility? [1 sentence with actual values.]

> 4. If we increase the time costs per child $c'$ from 10 to 15 hours per
>    week, what happens to fertility [1 sentence with actual values.]

> 5. For which of the above does the direction of the effect
>    depends on initial parameter values, and for which do you think the
>    direction of the effect is unambiguous? (Hint: you can verify the
>    analytical answers given in the previous section by
>    experimentation.) [A sentence for each parameter.]

> 6. Now change the initial parameter value of $c$ from 10 hours per
>    child to 2 hours per child before increasing wages from $15 per
>    hour to $30 per hour.

>   What is the effect of a wage increase on fertility now?
>   [1 sentence with actual values.]

>   Describe the relative size of the income and substitution effects
>   that must be true for this effect to hold. [1 sentence saying which
>   effect must dominate, no numbers needed.]

> 7. Review of concepts. For each of the following scenarios, specify
>   how you could operationalize the scenario by changing a parameter,
>   and what effect the change in the parameter would be expected to
>   have on fertility. (For example, if the scenario was a cut in the
>   inheritance tax, you could say that on average this would increase
>   non-wage income I, and that the effect of increasing I would be
>   unambigously to increase fertility.)

> A. The government reduces the costs of elementary schooling (e.g. by
>    getting rid of school fees). What parameter would you change, and
>    what effect would this have on fertility?


> B. Male income falls, but we live in a society in which only women
>    take care of children. What parameter would you change, and what
>    effect would this have on fertility?


> C. Men becoming more involved in child care. One way of
>    operationalizing is to increase $T$, the time that could be spent
>    working or taking care of children. What other parameter would one
>    want to change? How could these changes be made in a way that had
>    no effect on fertility?

> D. Infant mortality falls. If we conceptualize the costs per child as
>    costs per surviving child, then this effectively lowers the costs of
>    children. What parameters would you change, and what effect would
>    these changes have on fertility?


# Part 2. Fertility Decline in Switzerland

In this homework, we take advantage of the wonderful coincidence that
R has a built in data set from the literature on the historical
demographic transition. The dataset has fertility, mortality, and a
number of economic and social indicators for 47 French-speaking
provinces of Switzerland in the years around 1888.

The goal of this homework is to see empirically what the determinants
of lower fertility appear to be -- and to see which of our economic
theories seem to be applicable in this case.


## 2.1 Understanding the variables

Read results of help(swiss) and answer the following T/F questions
```{r}
help(swiss)


```

The fertility measure goes from 0 to 100, with 100 being an estimate
of the biological maximum fertility (under normal conditions). If
you're interested in the details, see
(http://opr.princeton.edu/archive/pefp/indices.aspx).


Q1.1 Agriculture is the percentage of land area  dedicated to farming.
A. True
B. False
{{
qnumber = 1.1
correct.answer = "B"
hint = "Ask someone for help if you can't display the help page for
the swiss dataset. It should appear using the 'help(swiss)' code chunk
given above."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "It's percent of males involved in agriculture as occupation."
}}


Q1.2 Education and examination are measured only for males
A. True
B. False
{{
qnumber = 1.2
correct.answer = "A"
hint = "Draftees are all male"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "True, but higher male education is probably associated
with higher female education -- so it is probably not incorrect to
think of this as a proxy for the education of both parents and of
children of both sexes."
}}

Q1.3 No measure of income is available
A. True
B. False
{{
qnumber = 1.3
correct.answer = "A"
hint = "Sorry, no clever hint."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Yes, a common problem with historical analysis."
}}


Q1.4 "Catholic" is the fraction, not percentage, of people who are
Catholic.
A. True
B. False
{{
qnumber = 1.4
correct.answer = "B"
hint = "Try 'head(swiss)' to look at the data; or read the help page."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "True, but not particularly important."
}}



Q1.5 What percent of children die before the age of 1 in Moutier?
```{r}
swiss["Moutier","Infant.Mortality"]
```
A. 0.203
B. 20.3
C. 2.03
{{
qnumber = 1.5
correct.answer = "B"
hint = "The question is being very literal: 'what percent'?"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "A lot of children died in their first year -- and even
more before they turned age 5."
}}



Q1.6 What is the district with the highest examination scores?
```{r}
swiss[which.max(swiss$Examination),]
```
A. The Valley of Geneva
B. Bern
C. Some very rural place
{{
qnumber = 1.6
correct.answer = "A"
hint = "V. stands for Valley."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Geneva is also presumably the most industrially developed."
}}



# 2. Theoretical Expectations

Before going further on the analysis, write down what you expect the
relationship between fertility and the other variables to be according
to Becker's quantity-quality model.

Here is a model answer to show the format to use for the
other parts.

Infant mortality

(Answer: Higher infant mortality would effectively increase the price
of child quality. This would encourage parents to instead invest in
quantity, raising fertility. So I would expect higher fertility to be
found in provinces with higher infant mortality.)


## Graded Questions on Expected Effects in Switzerland

> 8. Catholic
> (Hint: Assume that this raises the psychological cost of limiting fertility.)

> 9. Education
> (Hint: Give a two-part answer: first, the effect on parents
> having more education; and second, the effect children having more
> education)

> 10. Examination
> (Hint: Think of this as a measure of the result of more investments in
> the "quality" of children)

> 11. Agriculture



# 3. Analysis


What is the observed relationship between mortality and fertility?


Plot it, using fertility as the outcome on the Y-axis
```{r}
plot(Fertility ~ Infant.Mortality, data = swiss)
abline(lm(Fertility ~ Infant.Mortality, data = swiss), lty = 2)
```


Q3.1 Higher infant mortality is associated with

A. Higher fertility

B. Lower fertility

C. No clear pattern in fertility

{{
qnumber = 3.1
correct.answer = "A"
hint = "Consult your plot and the fitted line."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "This is consistent with our expectations above."
}}


2. What is the relationship between "Agriculture" and fertility?

```{r}
## Cut and paste and adapt the code above for Infant.Mortality above
```

Q3.2 Higher proportions working in agriculture are  associated with

A. Higher fertility

B. Lower fertility

C. No clear pattern in fertility

{{
qnumber = 3.2
correct.answer = "A"
hint = "Consult your plot and the fitted line."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Is this consistent with your expectations above?"
}}


3. What is the relationship between "Examination" and fertility?

```{r}
## Cut and paste and adapt the code above for Infant.Mortality above
```

Q3.3 Higher examination scores are  associated with

A. Higher fertility

B. Lower fertility

C. No clear pattern in fertility

{{
qnumber = 3.3
correct.answer = "B"
hint = "Consult your plot and the fitted line."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Is this consistent with your expectations above?"
}}


4. What is the relationship between "Education" and fertility?

```{r}
## Cut and paste and adapt the code above for Infant.Mortality above
```

Q3.4 More education is associated with

A. Higher fertility

B. Lower fertility

C. No clear pattern in fertility

{{
qnumber = 3.4
correct.answer = "B"
hint = "Consult your plot and the fitted line."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Is this consistent with your expectations above?"
}}


5. What is the relationship between "Catholic" and fertility?


```{r}
## Cut and paste and adapt the code above for Infant.Mortality above
```

Q3.5 More Catholic provinces tend to have

A. Higher fertility

B. Lower fertility

C. No clear pattern in fertility


{{
qnumber = 3.5
correct.answer = "A"
hint = "Consult your plot and the fitted line."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Is this consistent with your expectations above?"
}}



## Graded Questions on Observed and Expected Effects in Switzerland


> 12. Write a paragraph [< 100 words] summarizing which features (if
> any) of the Swiss data were consistent with your expectations from our
> micro-economic models and which features (if any) were
> inconsistent. Feel free to provide any further interpretation.


Congratulations! You have completed the lab for week 8.




