---
title: "The Bet: Over and over again"
output: html_notebook
---


Julian Simon liked to proclaim that

> The only true measure of scarcity is price ...

In this lab we are going to take advantage of a database put together by David Jacks. The data for his paper "From Boom to Bust: A Typology of Real Commodity Prices in the Long Run" (2013, NBER Working Paper 18874) has consistent series of commodity prices for everything ranging from barley to zinc. For our lab we've selected prices since 1900.

We're going to ask the following:

1. What has been the long-term trend in commodity prices?

2. Was Simon lucky to have won the bet?

3. Do commodity price trends differ by commodity in a way that we can
   understand?

4. Finally, we will use our Malthusian "harvest" model to learn about
   renewable resources.


```{r}
tot = 0
source("http://courses.demog.berkeley.edu/goldstein175/Basket/check_thebet.R")
source("http://courses.demog.berkeley.edu/goldstein175/Basket/check_functions.R")
```


# Part 1. Read in data and graph trends in commodity prices

These commodity prices are indexed in two ways. First, the nominal
prices have been adjusted for inflation. Second, all prices are
indexed to 100 in 1900. This makes it easy for us to compare across
commodities and over time.

```{r}
## get commodity data from class website
path <- "http://courses.demog.berkeley.edu/goldstein175/Basket/"
## path <- "../Basket/"
file <- "Real_commodity_prices_1900_2015.csv"
url <- paste0(path, file)               # full url
commod <- read.table(url, sep = ",") # read the data into, with column header
head(commod) # looks at first few lines
```
You can scroll to the right and left using the black arrows in the display of the data.

Q1.1 What does the number 110.51 under Zinc mean?

A.  Zinc costs 110.51 cents per pound

B.  Zinc prices increased by about 110 percent between 1900 and 1904

C.  Zinc cost about 10.51 percent more in *nominal* terms in 1904 than in 1900

D.  Zinc cost 10.51 percent more in *real* terms in 1904 than in 1900

{{
qnumber = 1.1
correct.answer = "D"
hint = "Prices are indexed to 1900 and are adjusted for inflation"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Real prices adjust for inflation"
}}


Q1.2 What is the major use of Chromium?

A.  For browsing the internet

B.  As Chrome plating and stainless steel

C.  As vitamin supplement for breakfast cereal

D.  As a precious metal for storing value

{{
qnumber = 1.2
correct.answer = "B"
hint = "https://en.wikipedia.org/wiki/Chromium"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "According to wikipedia, 85 percent is apparently used
for this one purpose. Are there substitutes for protecting steel?"
}}

To select a value from this matrix of data you can index by the label
of the row and column. For example,
```{r}
commod["1963", "Chromium"]
```
tells us that the real price of Chromium in 1963 was 55 percent of that
in 1900.

> (Note: in the rest of the lab, unless we say "nominal price", we mean
> "real price" when we use the word "price".)

# Part 2. Visualizing commodity prices

Let's see how the price of Chromium varies over time.

```{r}
year.vec <- rownames(commod)
plot(year.vec, commod$Tin, type = "l")
```
Q2.1 How would you describe the history of Chromium prices?

A. Steadily rising

B. Steadily falling

C. Very volitile, with little clear long-term trend

D. Very volitile, with a clear upward trend

{{
qnumber = 2.1
correct.answer = "C"
hint = "Sorry, no hint"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "There's clearly a lot of volitility. Although the price in 2015 is higher than in 1900, this doesn't look like a clear trend, since for much of the 20th century the price was lower than in 1900."
}}

## Repeating for Tin

Plot the price of tin. You can use the same code we used for Chromium, modifying just slightly.
```{r}
## put your code for Tin here
```
Let's add a vertical line at 1980 when Ehrlich chose Tin.
```{r}
abline(v = 1980)
```
Q2.2 Why do you think Ehrlich chose Tin as one of his commodities to bet on?

A.  Tin is difficult to substitute for

B.  Tin had shown a sharp increase in price for the decade or so before 1980 and so it looked like it would continue.

C.  The price had been going down before 1980 and was ready for a recovery

D.  There was evidence that the tin mines were running out of ore

{{
qnumber = 2.2
correct.answer = "B"
hint = "All are reasonable, except C which is incorrect. Try again if you didn't answer B."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Apparently, tin prices were controlled in part by a cartel of producers. See https://www.itri.co.uk/index.php?option=com_mtree&task=att_download&link_id=49605&cf_id=24 for a discussion from an industry research source."
}}


# Part 3. Replicating actual bet

In the original bet, Ehrlich was allowed to choose any commodities he
wished. He chose Chromium, Tin, Tungsten, and Nickel. In our data, we
don't have Tungsten. We can check to see if Simon still wins with the
three commodities we have.

We will use a function to compare commodities over time. This uses the
indexing by label we did above. It returns the change in value ending
year of a $1000 purchase of the commodity basket in the start year. So
a value of +100, would mean that the value of commodity basket
increased by $100.

Q3.1 If the value of the basket increased by $100, who would win?

A. Simon

B. Ehrlich

C. Neither, because the value of the basket doesn't tell us if
*all* of the commodities increased in price at the same time.

{{
qnumber = 3.1
correct.answer = "B"
hint = "Prices went up"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "An increase in prices suggests an increase in
'scarcity' (according to Simon)."
}}


```{r}
bet.fun <- function(start, end, basket.vec, data)
{
    ## start = 1980
    ## end = 1990
    ## data = commod
    ## basket.vec <- c("Copper", "Chromium", "Nickel", "Tin")
    year <- rownames(data)
    start.price <- data[year == start, basket.vec]
    end.price   <- data[year == end, basket.vec]
    ## we invest 1000, buyin an equal $ amount in each commodity.
    start.funds <- 1000
    dollars.per.commod <- start.funds/length(basket.vec) # e.g. $250 with 4
    ## amount purchased of each commodity
    holdings.vec <- dollars.per.commod/start.price
    ## value at the end
    value.at.end <- sum(holdings.vec * end.price)
    change.in.value <- value.at.end - start.funds
    return(change.in.value)
}

who.wins.fun <- function(change.in.value)
{
    ifelse(change.in.value > 0, "ehrlich", "simon")
}

```

Let's try this for 1980 to 1990 with Ehrlich's basket.

```{r}
ehrlich.basket.vec <- c("Copper", "Chromium", "Nickel", "Tin")
original.bet.result <- bet.fun("start" = 1980,
                      "end" = 1990,
                      "basket.vec" = ehrlich.basket.vec,
                      "data" = commod)
who.wins.fun(original.bet.result)
```

Q3.2 How much would Simon have won with our version of the "bet"?

A. About 300 dollars

B. He wouldn't have won. He would have lost about 300 dollars

C. About 500 dollars

D. About 100 dollars

{{
qnumber = 3.2
correct.answer = "A"
hint = "Insert a line: 'print(original.bet.result)' in the code chunk to see the result of the bet."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "A negative number means prices went down, and so Simon would have won."
}}




# Part 4. Does Simon win in other periods?

Commodities are highly volitile. Even if commodities were generally
rising, Ehrlich could have been unlucky about the timing of his bet.

Let's check over all years since 1900. We can construct a "loop" using
R that will repeat the bet according to our instructions. We can then
save the output in a vector and make a judgement about whether Ehrlich
was was simply unlucky.

Here's a simple example of a loop, in which we repeat an action 7 times,
incrementing the value of "i", and printing this value along with a label
"iteration".
```{r}
for (i in 1:7)   ## the variable "i" takes the value 1, 2, 3, ...
{
    ## anything in between {} is done once
    ## for each value that "i" takes
    print(c("iteration", i))
}
```

Now let's  a loop to see who won the bet in every year of the last
century.

```{r}
start.vec <- 1900:2005        ## a vector of years to start the bet
bet.result.vec <- rep(NA, length(start.vec)) ## an empty vector to be
                                             ## used to store the
                                             ## results
names(bet.result.vec) <- start.vec ## labeling the elements (you can ignore)
for (i in 1:length(start.vec))
{
    bet.result.vec[i] <- bet.fun(start = start.vec[i],
                                 end = start.vec[i] + 10,
                                 basket.vec = ehrlich.basket.vec,
                                 data = commod)
}
print(bet.result.vec)
```

To summarize our results, let's look at them in several ways.

Plot the numbers
```{r}
plot(start.vec, bet.result.vec)
abline(h = 0)
text(1960, 1500, "Ehrlich wins", col = "blue")
text(1960, -250, "Simon wins", col = "red")
```
In the graph, it looks like Ehrlich would have won more of the time. But not a lot more.

Let's tabulate how many times Ehrlich and Simon would have won:
```{r}
## Convert to a vector of "simon" and "ehrlich", depending on who won
winners.vec <- who.wins.fun(bet.result.vec)
print(winners.vec)
```
```{r}
## count number of times each won using the "table()" function
table(winners.vec)
```

Q4.1 What percent of the time would Ehrlich have won of the 105 simulated bets from 1900 to 2005?

A. About 62 percent

B. About 58 percent

{{
qnumber = 2.4
correct.answer = "B"
hint = "Insert there are 105 bets, not 100."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "62/105 = 0.58."
}}


Make a fancier plot. (You don't need to understand this code.)
```{r}
plot(start.vec, bet.result.vec, type = "n") #an empty plot
abline(h = 0)
e <- which(winners.vec == "ehrlich")    # T or F, to subset text()
s <- which(winners.vec == "simon")
text(x = start.vec[e], y = bet.result.vec[e], labels = "e", col = "blue")
text(x = start.vec[s], y = bet.result.vec[s], labels = "s", col = "red")
text(1960, 1500, "Ehrlich wins", col = "blue")
text(1960, -250, "Simon wins", col = "red")
```


# 5. Generalizing the bet to other commodities

A change in the supply of a particular commodity may or may not cause
a large price change -- depending on the availability of
substitutes. (Note: the results from this section will be useful for
your write-up of the graded questions at the end of the lab.)

Q5.1 Pick a commodity that you think has a lot of substitutes and see
     what fraction of the all of the 10 year bets from 1900 to 2005
     Ehrlich would have won.

Q5.2 Repeat Q4.1 for a commodity that you think is difficult to
     substitute for.

The following code provides a template for a single commodity, in this
case "Chromium".

```{r}
start.vec <- 1900:2005
result.vec <- rep(NA, length(start.vec))
for (i in 1:length(start.vec))
  {
  result.vec[i] <- bet.fun(start = start.vec[i],
                      end = start.vec[i] + 10,
                      basket.vec = c("Chromium"),
                      data = commod)
}
winner.vec <- who.wins.fun(result.vec)
cat("counts:\n")
table(winner.vec)
cat("proportions:\n")
prop.table(table(winner.vec))
```

Well done. You are finished with the computing portion of Lab 4.

# Part 5. Graded Questions


Reminder: These questions should be answered at bCourses. (If the quiz
"forms" are not available yet, write down your answers in another file
to be cut-and-pasted later, when the forms become available.)
```{r}
browseURL("https://bcourses.berkeley.edu/courses/1459366/quizzes")
```



> 1. Based on your analysis above, did commodity prices increase
>     steadily with population growth over the course of the 20th
>    century? [A few sentences, carefully worded, are fine here.]

> 2. Was Simon simply "lucky" or "right", or both?
>   [Explain your answer in a 2 or 3 sentences.]

> 3. Discuss the results of your analysis of an easily substitutable
>   commodity.

>   i. Tell why you think that this commodity would be easily
>   substitutable, explaining what "service" it provides. [1 or 2
>   sentences]

>  ii. What did your analysis tell you about whether prices
>  consistently increased? [1 or 2 sentences]

>  iii. Interpret your results [1 or  2 sentences]

> 4. Repeat question 3 (i,ii, and ii), but for the commodity you thought
>   would be hard to substitute for.

> 5. Did your results in 3 and 4 differ substantially from each other or
>    from the results of Ehrlich and Simon's bet? [< 50 words]

> 6. What are two reasons prices might not reflect the long-term
>    availability of a commodity. [A sentence on each.]

> 7. Renewable resources

> i. Imagine a community living by a large lake full of fish. Each
>    household in the community catches ten fish per day, which is the
>    maximum sustainable yield. A technological innovation allows the
>    households to increase their catch to twelve fish per day. Describe
>    will happen to their yields in the short and longer term. [ < 50 words]

> ii.	How does this relate to the idea of the tragedy of commons?
> 	[A sentence or two is fine.]

Congratulations! You are finished with Lab 4.




