---
title: "Econ/Demog C175 Lab 2: Solow model analysis"
output: html_notebook
---

```{r}
tot = 0
source("http://courses.demog.berkeley.edu/goldstein175/Basket/check_solow.R")
source("http://courses.demog.berkeley.edu/goldstein175/Basket/check_functions.R")
```

# Overview

In this lab we will work with the neo-classical growth model developed
by Solow and Swann. Our goals are:

1. To derive some analytical properties of the model, assuming a
Cobb-Douglas production function.

2. To use understand the dynamics of income and savings using a
simulation app.

3. To understand how the capital intensification may or may not lead
   to increasing inequality, depending on the nature of the production
   function.

The order of this lab will be slightly different in that we will start
with some mathematical derivations and then go on to do some computing
afterwards.

# 1. Introduction

In the Solow model, the capital per worker and output per worker are
in their steady state when investment exactly balances population
growth and depreciation. This is when
$$
s y(k) = (n+d) k.
$$

# 2. Analytical questions

Note: In this lab, we are beginning with some analytic, pencil and
paper problems. Please answer them on bCourses.

1. Let the production function be Cobb-Douglas, with $y(k) =
k^(.4)$. Assume that the savings is 20 percent, population growth is 1
percent, and depreciation is 4 percent.

(Express your answer as a general formula in terms of the symbols "s",
"k", "n", "d", and the exponent "\alpha" of the Cobb-Douglas production
function.

For example, for the steady-state level of capital per worker, we
substitute the production function y = k^a into the steady-state
condition above to give
$$
s y(k) = s k^\alph = (n+d) k
$$
Re-arranging, then gives us a formula for k in the steady state:
$$
( s / (n + d) )^(1 / (1-\alpha) ) = k
$$

You can then substitute the values given in the problem and provide
your numerical answer.)

a) What is the steady-state level of capital per worker?

b) What is the steady-state of output per worker?

c) What is the steady-state level of consumption per worker?

2. Now assume population growth is instead -0.5 % (approximately the growth
   rate when every couple has 1.7 children), but that all other
   parameters stay the same.

a) By what percent does the steady-state output per worker change?

b) How much higher would the savings rate need to be to keep the steady-state
capital/labor ratio (and output) at the same levels as when population
growth was 1%?

c) How much lower would steady-state consumption be? (Express your
answer in percentage terms, e.g., "consumption would decline by 50%"
(not the real answer))


# 3. Questions with the App

For the following question, you may find it useful to experiment with
the "Solow_2017" app available at

http://shiny.demog.berkeley.edu/josh/solow_2017/

(You can also use the app to get a rough check your answers to the
analytical questions above.)

1. Assume a technological innovation, like the availability of
   electricity, increases output per person at all levels of the
   capital/labor ratio from a previous level of 1.0 to a new level
   of 1.3.

a) Describe in words what happens in the short run to output. (E.g.,
   how large is the immediate increase in output?)

b) Describe in words what happens in the long-run. (If the long-run
   steady state is higher than the short-term level of output right
   after the technology change, what is causing this additional
   increase?)

# 4. Growth and Inequality


In this part of the lab we will use R to calculate the shares of
income from capital and from labor from two production functions.

## i. Cobb-Douglas

Here we use a production function of the form
$$
y(k) = k^\alpha
$$


```{r}
k <- seq(1, 20, .1)
alpha <- 1/3
y.of.k <- k^alpha
## numerical slope = rises/runs
mp.k.numeric = diff(y.of.k)/diff(k)
plot(k[-1], mp.k.numeric)
## analtyic derivative, taking derivatitive of k^alpha with respect to k.
## (don't worry about this if you haven't had calculus)
mp.k.analytic = alpha * k^(alpha - 1)
lines(k, mp.k.analytic, col = "red", lwd = 2)
## we see they match quite well
```

Q1.1 Is the marginal product on capital
A. Constant with increases in capital
B. Falling  with increases in capital
C. Declining with increases in capital

{{
qnumber = 1.1
correct.answer = "C"
hint = "The marginal product on capital is the additional output from
an increment of capital, the slope of the production function with
respect to k."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "The marginal product on capital is equal to the slope of
the production function, which our plot shows is declining over
time."
}}

Now let's see what happens to capital's share of total output. We
assume here that the rate of return is equal to its marginal
product. So as we increase the capital per person, we will have two
counter-vailing forces: the amount of capital will increase, but the
rate of return will decrease. Let's see what effect dominates or if
the two effects cancel each other out.

```{r}
mp.k <- mp.k.analytic
output.from.capital.per.worker <- mp.k * k
total.output.per.worker <- y.of.k
output.from.labor.per.worker <- y.of.k - mp.k * k
capital.share <- mp.k * k / y.of.k
```

Now let's see what happens to capital share as we increase k
```{r}
plot(k, capital.share)
```

Q1.2 Does the capital share of income
A. Rise with increases in capital
B. Stay constant with increases in capital
D. Fall with increases in capital


{{
qnumber = 1.2
correct.answer = "B"
hint = "See our graph."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "With Cobb-Douglas production the rate of increase in
capital is exactly balanced with the decline in the rate of return,
keeping the capital share of income constant."
}}


## ii. An alternative production function

Let's modify the Cobb-Douglas production function slightly so that it
is
$$
y(k) = k^\alpha + k/10
$$

Modify the code below to work with this new production
function. (Hint: the derivative with respect to k of k^alpha + k/10 is
alpha * k^alpha + 1/10.)


```{r}
k <- seq(1, 20, .1)
alpha <- 1/3
y.of.k <- k^alpha ### <--- modify this line.
## numerical slope = rises/runs
mp.k.numeric = diff(y.of.k)/diff(k)
plot(k[-1], mp.k.numeric)
## analtyic derivative, taking derivatitive of k^alpha with respect to k.
## (don't worry about this if you haven't had calculus)
mp.k.analytic = alpha * k^(alpha - 1) ### <--- modify this line, too!
lines(k, mp.k.analytic, col = "red", lwd = 2)
## you should see that they match quite well
```

Q2.1 Is the marginal product on capital still declining
A. Yes, but perhaps less quickly
B. No, it is no longer declining
{{
qnumber = 2.1
correct.answer = "A"
hint = "Consult your graph to see if the marginal product is declining."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Yes, it is still declining. If you compare with
Cobb-Douglas, the rate of return looks like it declines more slowly,
especially for large values of k."
}}


Now let's see what happens to capital's share of total output.

```{r}
mp.k <- mp.k.analytic
output.from.capital.per.worker <- mp.k * k
total.output.per.worker <- y.of.k
output.from.labor.per.worker <- y.of.k - mp.k * k
capital.share <- mp.k * k / y.of.k
## graph our result to see what happens to capital share as we increase k
plot(k, capital.share)
```

Q2.2 Does the capital share of income
A. Rise with increases in capital
B. Stay constant with increases in capital
D. Fall with increases in capital

{{
qnumber = 2.2
correct.answer = "A"
hint = "See our graph."
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "We now see that the more capital we have per worker,
the greater the capital share of the economy is! This is what Piketty
argues is likely to happen. As growth rates fall, capital per worker
will increase, and with it an increase in the capital share of the
economic output."
}}


# Graded questions


1. (see beginning of lab)

2. (see beginning of lab)

3. What level of population growth would maximize income per capita? (Hint:
   you don't need calculus for this problem. Try thinking about it
   with a diagram.)

4. Many are worried about the effects of robots on inequality. But
   perhaps robots will also depreciate more quickly than earlier forms
   of productive capital.

a) What would happen to capital per worker if we invested our savings
   in fast depreciating robots instead of slower depreciating
   traditional productive capital? (Please make the (unrealistic?)
   assumption that per dollar robots have the same effect on
   production as traditional capital.)

b) Assume, as per Piketty (and per our 2nd example of a production
   function above), that the capital share of output increases with
   capital intensification and decreases when capital per worker
   declines. Would our faster depreciation increase inequality or
   reduce it?

Congratulations! You are finished with lab 3.
