---
title: "Econ/Demog C175 Lab for Week 9: Fertility and the Great
Recession in the United States
output: html_notebook
---

Note: this is an slightly adapted version of our in-class lab. There
are no auto-checked answers of the multiple choice questions. The
graded questions are at the end.

# The Great Recession and Fertility in the United States

To begin with let's look at the plot of historical fertility rates in
the United States.

Q1. Does a better economy ...

A. Always lead to higher fertility

B. Sometimes lead to higher fertility and sometimes lead to lower
fertility

C. Usually lead to higher fertility

D. I see no relationship?


Q2. In terms of our cost-of-time model, if a better economy raised
fertility, then this would mean

A. The substitution effect dominates the income effect

B. The income effect dominates the substitution effect

Q3. If men and women become more alike, then what would we expect?

A. Stronger substitution effects

B. Stronger income effects

C. It's not so clear ...

Q4. What do we expect in the Great Recession?

A. Lower fertility

B. Higher fertility

C. No change

Q5. What do we expect now that the Great Recession is over?

A. Fertility to return to it's previous level?

B. Fertility to be even higher, making up for lost time.

C. Fertility to stay lower, until people are really sure the recovery
will continue.

# Data analysis

We will look at unemployment rates and fertility rates.

## Importing the data

```{r}
prefix <- "http://courses.demog.berkeley.edu/goldstein175/Basket/"
usa.df <- read.csv(paste0(prefix,"usa_tfr_ue_by_year.csv"))
state.df <- read.csv(paste0(prefix,"state_tfr_ue_by_year.csv"))
```

## The Great Recession

### Nationally

```{r}

## save ourselves some typing
tfr <- usa.df$tfr
ue <- usa.df$ue
year <- usa.df$year                     # 2001:2015
## to compare UE in year of conception
ue.before <- ue[year %in% 2001:2014]
tfr.after <- tfr[year %in% 2002:2015]
```

We can plot the relationships
```{r}
par(mfrow = c(2,2))
plot(year, tfr, type = "o",
     col = "blue",
     main = "Total Fertility Rates, USA")
plot(year, ue, type = "o",
     col = "red",
     main = "Unemployment, USA")
plot(ue, tfr, type = "p",
     col = "black",
     main = "TFR by unemployment, USA")
abline(lm(tfr ~ ue))
plot(ue.before, tfr.after,
     main = "TFR by prior year's unemployment")
abline(lm(tfr.after ~ ue.before))
```

Q. Is higher unemployment associated with lower fertility?

A. Yes, throughout the whole period

B. No, higher unemployment is associated with higher fertility.

C. It depends on which period we look at

D. I have no idea!

Q. Does it make a difference if we look at conditions at the time of
conception vs. time of birth?

A. Not much


Let's look at the period up to 2011, separately.

```{r}

s <- year %in% 2001:2011
par(mfrow = c(2,2))
plot(year, tfr, type = "o",
     col = "blue",
     main = "Total Fertility Rates, USA")
points(year[s], tfr[s], pch = 19, col = "grey")
plot(year, ue, type = "o",
     col = "red",
     main = "Unemployment, USA")
points(year[s], ue[s], pch = 19, col = "grey")
plot(ue, tfr, type = "p",
     col = "black",
     main = "TFR by unemployment, USA")
points(ue[s], tfr[s], pch = 19, col = "grey")
## abline(lm(tfr ~ ue))
abline(lm(tfr.after ~ ue.before, subset = s),
       col = "grey", lty = 2)
plot(ue.before, tfr.after,
     main = "TFR by prior year's unemployment")
## abline(lm(tfr.after ~ ue.before))
abline(lm(tfr.after ~ ue.before, subset = s),
              col = "grey", lty = 2)
points(ue.before[s], tfr.after[s], pch = 19, col = "grey")
```

So, we see that the recession seems to have lowered fertility on a
national level.

We can also look at the states to see if states that were harder hit,
saw a bigger decline in fertility.

Import a fancy function to plot an panel of states in the shape of the USA
```{r}
source(paste0(prefix,"recession_functions_for_class.R"))
```

Now display the state TFR and UE.

```{r, fig.height=8, fig.width=10}

state.panel.plot(x = state.df$year,
                 y = state.df$tfr,
                 state = state.df$state.name,
                 vline = 2008,
                 hline = 2.0,
                 new = T,
                 type = "l",
                 col = "grey",
                 legend.ylab = "tfr",
                 legend.xlab = "year",
                 title = "TFR by state, 2001-2015, all",
                 ylim = c(1.5, 3))
state.panel.plot(x = state.df$year,
                 y = state.df$tfr.filt,
                 state = state.df$state.name,
                 vline = 2008,
                 hline = 2.0,
                 col = "blue",
                 type = "l",
                 lwd = 3,
                 new = F,
                 ylim = c(1.5, 3))

```

Now look at UE

```{r, fig.height=8, fig.width=10}
## now do unemployment
state.panel.plot(x = state.df$year,
                 y = state.df$ue,
                 state = state.df$state.name,
                 vline = 2008,
                 hline = c(.05, .1),
                 col = "red",
                 type = "l",
                 lwd = 2,
                 new = T,
                 title = "Unemployment by state, 2001-2015, all",
                 ylim = c(0, .15))

```


```{r}
## now do a scatter plot
par(c(1,1))
plot(state.df$ue,
     state.df$tfr.filt, type = "p")
m0 <- lm(tfr.filt ~ ue , data = state.df)
abline(m0)
```

Here are some (optional) questions to consider:

Q. Do "oil and gas" states (e.g. the Dakotas) see less effect of Recession?

Q. Does their fertility also not fall as much?

Q. How about coastal states?

Q. Which states have biggest increase in unemployment?

Q. Which states have biggest decline in fertility?


# Graded Questions

We asked in class why fertility hasn't recovered as unemployment has
gone back down to pre-recession levels. One explanation is that
people's current circumstances have recovered, but that they are now
more uncertain (than before the recession) about their
futures. The fact that the Great Recession happened at all has made
people wary that something else bad will happen again.

We don't have measures of attitudes in the American Community Survey
that we are using to measure fertility and economic
conditions. However, we can disaggregate the population into those
with more and less education. If we assume that those with less
education would be more at risk in the event of another economic
downturn, then perhaps the less educated has also shown less recovery
in fertility.

In this last section, you will analyze the fertility of women with and
without a college diploma.

## Importing the data

```{r}
df <- read.csv(paste0(prefix,"educ_tfr_unemp_by_year.csv"))
tfr.college <- df$tfr[df$college == TRUE]
tfr.nocollege <- df$tfr[df$college == FALSE]
ue.college <- df$ue[df$college == TRUE]
ue.nocollege <- df$ue[df$college == FALSE]
year.vec <- 2001:2015
```

## Looking at the recovery



Graded Question 1. Submit a plot of the trends in TFR for those with a
college education.

```{r}
## Here is some code for code for plotting the unemployment rates
## by education. You can modify this slightly to produce a plot of tfr rather than "ue" (unemployment).
plot(year.vec, ue.nocollege, pch = "n", type = "o",
     col = "red", ylim = c(0, .2),
     ylab = "unemployment rate")
## the "lines()" functions adds a line to an existing plot
lines(year.vec, ue.college, pch = "c", type = "o",
     col = "blue")
```

```{r}
## You might want to plot the smoothed data
tfr.college.filt <- filter(tfr.college,
                             c(1/3,1/3,1/3))
## this gives a running three year average
```

Graded Question 2. From your plot, describe the differences between
the time trend in fertility by education. Is there a difference in
levels? Is there a difference in the time trend other than levels?
(Hint: the data for the more educated is more noisy because it has
fewer observations, so don't take the year-to-year fluctuations too
seriously. Instead, look at the general trend over time.)
[ A few sentences is fine < 50 words.]

Graded Question 3. Looking at the plot of unemployment above, were the
less educated hit harder by the recession? [ < 25 words]

Graded Question 4. Does it appear that the more and less educated
reacted to the Great Recession and its aftermath in different ways?
Does this lend support to the hypotheses that one of the reasons that
fertility hasn't recovered after the recession is that people are
still unsure if the recovery will last? (Hint: there is no one correct
answer here. Just give your reasoned interpretation.)[ < 50 words]





