---
title: "Econ/Demog C175 Lab 2: Malthus in Sweden"
output: html_notebook
---

```{r}
tot = 0
source("http://courses.demog.berkeley.edu/goldstein175/Basket/check_malthus.R")
source("http://courses.demog.berkeley.edu/goldstein175/Basket/check_functions.R")
```


# Overview

Sweden has the longest historical series of reliable vital statistics
in the world. The Human Mortality Database, founded at Berkeley, has
records for counts of births, deaths, and population size since 1751,
well before the demographic transition.


Malthus visited Norway and Sweden and describes his travels in his 2nd
edition of (An Essay on the Priciple of
Population). [http://www.econlib.org/library/Malthus/malPlong7.html#Bk.II.Ch.I]. Please
read these two chapters (Book II, Chapter I, "On the Checks to
Population in Norway" paragraphs II.I.1 through II.I.8; and Chapter
II, "On the Checks to Population in Sweden" paragraphs II.II.1 through
II.II.9) for a better understanding about Malthus's thinking on how
the environment and human custom interact to produce demographic
outcomes.


In this lab, our goals are to:

1. To assess if the 18th century Swedish series of births and deaths is
consistent with Malthusian theory and see if we can detect a weakening
of Malthusian regime in the 19th and 20th centuries.

2. To learn how to use custom functions like malthus.plot.fun() by
   changing their arguments (without rewriting the functions themselves).

3. To use the non-graded interactive questions we have written for you
(e.g., Q.1.1, Q.1.2).

Note: there are TWO types of questions in Lab 2. First, we have the
non-graded interactive questions in the special code chunks (which we
did not have in Lab 1). These you do in RStudio. The goal of these is
to help your learning of the material, not to assess you. Second, as
in Lab 1, we have questions at the end that will be graded. These you
do in bCourses.




# Part 0: Import the data


```{r}

## Note: you need to run this code to do the lab. But understanding it
## is optional.

## Read in the data from a file on the web using next two lines
url <- "http://courses.demog.berkeley.edu/goldstein175/Basket/SWE_pop_birth_death.txt"
df <- read.table(url, header =T)

## Display the first few rows of the object "df"
head(df)

## Create a separate object for popualtion size, births, deaths, and year.
## Note: "read.table()" imports as a "data.frame", which stores each
## column of a data set as a separate element, which can be accessed
## with the dollar-sign, as below.
N <- df$N                               # population size
B <- df$B                               # births
D <- df$D                               # deaths
year <- rownames(df)
## Note: N, B, D, and year are vectors. In the last lab we used ".vec" in the names of the vectors. Here we omit this to make our code more readable.

```

# Part 1: Inspect and understand the data

Plot the history of Swedish population size, counts of births, and counts of deaths.

```{r}
plot(year, N, type = "l")
```

Q1.1 How big is the population of Sweden in 1800

A.  about 20 thousand

B.  about 200 thousand

C.  about 2 million

D.  about 20 million

{{
qnumber = 1.1
correct.answer = "C"
hint = "Sweden has about 10 million people today"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "2e+06 means 2 * 10^6, which equals 2 million"
}}

To get a more readable plot we can change units
```{r}
million = 1000000
N.millions <- N/million
plot(year, N.millions, type = "l")
```

Q1.2 What is the average (exponential) growth rate of Sweden from 1751
to 2014.
```{r}
( log(N[year == "2014"] - log(N[year == "1751"]) )/ (2014 - 1751)
```
Which value is this closest to?

A.  1 percent

B.  6 percent

C.  0.5 percent

D.  0.005 percent.

{{
qnumber = 1.2
correct.answer = "C"
hint = "0.01 is 1 percent"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Yes. Incidentally, a growth rate of 0.5 percent has a
doubling time of about 140 years, which we can calculate from
$T_{doub} = log(2)/R$. So we should have seen roughly 2 doublings,
which checks with our graph of population history."
}}


```{r}
# plot of Deaths by year, with type = "line"
plot(x = year, y = D, type = "l")
```

Q1.3. In what year did Sweden have the most deaths?

A.  In about 2000, because the population age structure shifted to older ages.

B.  In about 1850

C.  In 1773

{{
qnumber = 1.3
correct.answer = "C"
hint = "Try an earlier date"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Amazingly even though the population was smaller, the rate of mortality was so high, that 1773 saw they highest number of deaths."
}}


Plot the births using code similar the code for deaths, which is given
below. (Hint: You just need to replace the "y" variable.)
```{r}
# This is the code for deaths by year (modify it slightly).
plot(x = year, y = D, type = "l")
```

Q1.4 When were births at their low?

A.  In the great depression (1930s)

B.  In the low fertility of the economic recession at the end of the 1990s.

C.  In about 1773

{{
qnumber = 1.4
correct.answer = "C"
hint = "Try an earlier date"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "Amazingly, the same date as peak mortality."
}}



# Part 2: Time trends in crude rates (per capita births and deaths)

Let's get the births and deaths on a per capita basis. Demographers call these crude rates. ("Crude" because we don't refine the population at risk -- for example, we include women and men in the denominator of the birth rate.)

```{r get.rates, fig.height=5}
cbr = B/N                               # crude birth rate
cdr = D/N                               # crude death rate
par(mfrow = c(2,1))                     # two panels in one plot
plot(year, cbr, type = "l", main = "Crude birth rates, Sweden")
plot(year, cdr, type = "l", main = "Crude death rates, Sweden")
```

Q2.1 The time trends in this plot are:

A.  Basically the same as in the counts

B.  Quite different from the counts

{{
qnumber = 2.1
correct.answer = "B"
hint = "Compare this plot of rates with the earlier plot of counts"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "The trends in rates show long term declines, particularly after 1850. The trends in counts are confounded by the population increase."
}}

Q2.2. In what year was the peak death rate

A.  In the same year as the peak number of births

B.  In the the most recent year

C.  In 1773, the same year as the peak number of deaths

{{
qnumber = 2.2
correct.answer = "C"
hint = "Look for the spike in your graph of rates?"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "The spike in mortality in 1773 was very sharp."
}}

Q2.3 Google "1773 Sweden". What happened in this year?

A.  There was a war

B.  There was a famine

C.  The borders of Sweden changed.

{{
qnumber = 2.3
correct.answer = "B"
hint = "There wasn't a war"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "The 1773 famine was a catastrophic event. An
inspiration for Malthus's thinking?."
}}


# Part 3: Detecting Malthusian dynamics

In class, we gave a two-panel graphical explanation of the Malthusian
model. In the upper panel, we showed  the
relationship between wages and birth and death rates. In the lower
panel, we showed the relationship between population size and
wages. Together, these two panels create the feedback loop for a very
low growth Malthusian economy and population.

Here, we will look at whether one part of the Malthusian world appears
to hold: that births and deaths go in opposite directions in good and
bad times. We don't have a measure of wages, so our inference is going
to be indirect, looking at the relationship between births and deaths
by year.

Q3.1. What relationship between births and deaths would be consistent
with a Malthusian economy

A.  As births go up, deaths go up

B.  As birth rates go up, deaths rates go up

C.  As births go down, deaths go up

D.  As birth rates go down, death rates go up.

{{
qnumber = 3.1
correct.answer = "D"
hint = "Think of rates, not counts. In a bad year what would happen?"
instructions = "Replace the NA with your answer (e.g., 'A' in quotes)"
explanation = "In a bad times, death rates would rise (the positive check) and people would avoid having children (the preventive check)."
}}

```{r}
malthus.plot.fun <- function(from.year, to.year)
{
  ## a scatter plot of crude birth and death rates
  ## note: this function assumes that cdr and cbr already exist.
  ## If you're having problems, try running the code chunk above
  ## that creates the cbr and cdr
  my.years <- from.year:to.year
  s <- year %in% my.years
  plot(cdr[s], cbr[s], type = "n",
       xlab = "Crude death rate", ylab = "Crude birth rate")
  abline(lm(cbr ~ cdr, subset = s))
  text(cdr[s], cbr[s], year[s], cex = .5)
  my.cor <- round(cor(cdr[s], cbr[s]), 2)
  title(paste("Birth and Death Rates in Sweden, from", from.year,
              "to", to.year, "\n Correlation:", my.cor))
}

## now let's plot the first 50 years
malthus.plot.fun(from.year = 1751, to.year = 1800)
```
Q3.2 Which of the patterns in Q3.1 (A,B,C,or,D) does this match most closely?

{{
qnumber = 3.2
correct.answer = "D"
hint = "The downwards slope shows a negative relationship. Higher death rates are associated with lower birth rates."
instructions = "Replace the 'NA' with your answer (e.g., 'A' in quotes)"
explanation = "Higher mortality rates are in years with lower
fertility rates"
}}


Q3.3 Is the pattern seen from 1751 to 1800 consistent with Malthusian theory?

A. Yes

B. No

{{
qnumber = 3.3
correct.answer = "A"
hint = "See Q3.1 and Q3.2"
instructions = "Replace the 'NA' with your answer (e.g., 'A' in quotes)"
explanation = "Our data is consistent with Malthus's idea that in bad times birth rates fell and death rates rose, and that in good times death rates fell and birth rates rose."
}}

We see that birth rates are low when death rates are high (e.g.,
in 1772 and 1773). Even if we ignore these unusual years,
there is a generally negative relationship. Now plot the relationship
for the following time periods.

1801-1850
```{r}
## Modify the code we used for 1751 to 100 appropriately.
malthus.plot.fun(from.year = 1751, to.year = 1800)
```
1851-1900
```{r}
## Modify the code we used for 1751 to 100 appropriately.
malthus.plot.fun(from.year = 1751, to.year = 1800)
```
1901-2014
```{r}
## Modify the code we used for 1751 to 100 appropriately.
malthus.plot.fun(from.year = 1751, to.year = 1800)
```

# Part 4: Lab write up. Please limit each answer to 50 words or less.

> Remember: You need to answer these questions on bCourses.

> 1. Turn in your plot of the relationship between birth and death
> rates from 1801-1850.  Note: to save a
> graph, right click on it with your mouse and "Save image as ..."
>

> Completely optional: for those familiar with R, you can add your
> name to the plot using the text() function in R. Try adding it at
> the ncoordinates x = 1840, y = .02).

> 2. In which of the periods you explored above does the Malthusian
> relationship first break down? What is your basis for judging? What
> happens to the trend in birth and death rates that makes the
> Malthusian relationship go away?  [ < 50 words]

> 3 What does Malthus say is the main determinant of birth rates in
> *Norway*? (Hint: see II.I.3 and subsequent paragraphs in the
> reading).  Would this suggest that births would fall in the *same*
> year as deaths would rise, or a year or two after? [ < 50 words]

> 4 For the full Malthusian model, we would want to see if population
> size influences birth and death rates. How could you explore this
> graphically with these data? Describe the plot you would want to
> make. What would you put on the x-axis, and what on the y-axis,
> making sure to specify whether you mean rates and counts. What
> relationship would would support Malthusianism? [ < 50 words] (Note:
> you don't have to make this plot, just answer the above questions in
> words.)

> 5 One major innovation of the pre-modern period was the introduction
> of crops like the potato to Europe from the New World. In this
> problem, we will use the Malthus
> (app)[http://shiny.demog.berkeley.edu/josh/new_malthus] to
> investigate what effect Malthus would predict that this new staple
> crop for a country like Ireland. We will model this direclty by
> assuming that potatoes do not afffect fertility or mortality
> directly, but rather that they raise the effective wage rate from 1
> to 1.18 by making famers more efficient.

> 5.A Describe in words what happens in the *short term* as a result
> of the introduction of the potato. What happens to wages in the
> short term? What happens to birth and death rates in the short term?
> [< 50 words]

> 5.B According to Malthus, should the population size of Ireland
> return settle back to its inititial size long after the introduction
> of the potato? Why or why not? [< 50 words]

> 5.C In general, which parts of the model will change the wage level
> in the long-term, influencing individual welfare? For each of the
> following, indicate whether the long-term wage will be different
> than its initial value and
> explain in one sentence why or why not.
> 5.C.i An upward shift the b(w) function of births in terms of wages
> 5.C.ii An upward shift the d(w) function  of deaths in terms of wages
> 5.C.iii A larger initial population size N(0).
> 5.C.iv An upward shift in the function w(N) of wages in terms of population size.








