---
title: "A minimal example"
output: html_notebook
---

This is a minimal example for everyone to try to get started using R in our Rstudio Cloud environment inside a notebook. 


Let's calculate a fertility rate. Say there were 10 births to 45 women aged 20-25 in 1987.

```{r}
nBx = 10
nNx.female = 45
nFx = nBx/nNx.female
print(nFx)
```

R is a vectorized language. Here as an example using vectors, taken from Keyfitz and Flieger (1968, page 312) for 1851 France.

```{r}

## some data typed in by hand by Josh
million = 10^6 # to save us typing and to make code more readable
x <- c(15, 20, 25, 30, 35, 40) # ages at start of each interval, equivalent to 'seq(15, 40, 5)'
nKx.female.vec = c(1.55, 1.52, 1.44, 1.35, 1.27, 1.17) * million # suffix .vec just for us humans 
nBx.vec = c(39, 212, 283, 235, 147, 54) * 1000

## now calculate age-specific fertility rates
nFx.vec <- nBx.vec/nKx.female.vec   # can add, subtract, multiply and divide vectors 
## (Note this is element by element and not matrix multiplication, which has a different syntax)

## Plot the result
plot(x = x, y = nFx.vec, 
     type = "o",   # plotting type "o" means points _o_ver lines (try changing to "p", or "l", or "b")
     ylim = c(0, .25))
title("Period Age-specific Fertility in France, 1851")
```


Let's do one more example, calculating the TFR in France in 1851. The objects carry over from chunk to chunk. So we can just pick up where we left off ...

```{r}
n = rep(x = 5, times = length(x)) ## a vector of 5s of the right length
TFR = sum(n * nFx.vec) ## sum of the vector that results from the product of n * nFx.vec
print(TFR)
NRR = TFR * 0.57 * 0.4886
NRR
```


Some exercises:


1. Create a new chunk and use R to calculate the NRR using the approximation

$$
NRR = TFR \cdot \ell_\mu \cdot p_{female} 
$$


with the survival to the mean age of childbearing $\ell_\mu = 0.58$, the mean age of childbearing $\mu = 29.7$, and the proportion female at birth $p_{female} = 0.4886. (Note if you play with your mouse on top of the equation above -- or try maybe inserting a carriage return -- it should display nicely.)

2. Read the help page for 'plot' in the "Help" tab of the lower-right pane of your Rstudio session. See if you can re-do the graph of age-specific fertility in 1851 France, changing the color of the lines and points to "red" and changing the label on the x-axis to "Age at start of interval". (Hint: the command for how to change the labels of the axes is in the main part of the help page, but for the color you'll have to look under "Examples".)


Run the next chunk in order to import the necessary files for self-grading of the next multiple choice question.

```{r}
source("/BFDW/check_lab0_minimal.R")
source("/BFDW/check_functions.R")
```



3. Answer the following multiple choice question. If the NRR in France were found to be 0.97, does that mean that in 1851 

A. the intrinsic growth rate was less than zero

B. the observed population growth rate was less than zero

C. both of the above

{{
answer: yes, but this is the growth rate of the closed, stable population with these rates, not necessarily the observed growth rate of the open, actual population. In fact, according to K&F, the observed population had a rate of natural ecrease of about 0.5 percent.
hint: the intrinsic rate is what would be observed if rates were held constant for a long time in a closed (no-migration) population.
}}

