---
title: "Econ/Demog C175 Lab 4: Population Aging -- Are we having too
few children?"
output: html_notebook
---


In this lab, we will use Japan to illustrate the changes in
age-structure that are resulting from the demographic transition and
continued population aging.

Specifically, we will investigate:

1. What has happened to the share of elderly and children from
World War II until today.

2. What will happen if current levels of fertility and mortality
continue.

and

3. How to use stable age-structures and the Solow model to understand
the effect of fertility on consumption in the macro-economy.

> Note 1: This lab is somewhat longer than previous labs. Please plan
> accordingly. There is relatively little code for you to write but
> there is a substantial amount of material to cover.

> Note 2: Please be warned of the  varying notation for the
> exponential growth rate of the stable population. Economists use "n"
> and demographers use "r". We tend to mix this usage, in class and in this
> lab.



# Part 0. Preliminaries

We begin by downloading some data and functions to be used in the lab.


```{r}
tot = 0
source("http://courses.demog.berkeley.edu/goldstein175/Basket/check_aging.R")
source("http://courses.demog.berkeley.edu/goldstein175/Basket/check_functions.R")
```

> Note: the most helpful readings for this lab is Lee (2003) for Parts
> 1 and 2 and Lee et al (2014) for parts 3, 4, and 5.


# Part 1. Population change over the demographic transition.

The first declines in mortality and fertility happened gradually in
Europe. More recently, the pace of demographic transition has
sped up. An early example of rapid demographic change, for which we
have excellent statistics, is Japan.

Until shortly after World War II, Japanese fertility remained
high. Mortality rates were improving, and the resulting population
growth was very fast -- indeed, rapid population growth was one of
reasons that Japan gave for expanding its empire.

After the War, fertility began to decline very rapidly.

```{r}

tfr.data <- read.table("http://courses.demog.berkeley.edu/goldstein175/Basket/japan_tfr.txt")
year.vec <- tfr.data$Year
tfr.vec <- tfr.data$TFR
plot(year.vec, tfr.vec, type = "o",
     main = "Japanese Period Total Fertility Rate")
```

In less than a decade, fertility fell to replacement levels near two
children per woman.

Q1.1 What do you think happened in 1966?

A. There was a mistake in the data collection. Records were lost, or a
calculation error must have been made.

B. The economy crashed

C. 1966 was an inauspicious year according to the Japanese zodiac, the
year of the "fire horse" in which it was unlucky to have a daughter
born.

D. Everything in C *and*  further evidence of intentionality
is that births were slightly higher both before and after 1966.

```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer1.1 = NA
check(answer1.1)
```



We can see that the pace of mortality improvement was steady the
entire time.

```{r}
e0.data <- read.table("http://courses.demog.berkeley.edu/goldstein175/Basket/japan_e0_female.txt")
year.vec <- e0.data$Year
e0.vec <- e0.data$Female
plot(year.vec, e0.vec, type = "o",
     main = "Japanese Period Life Expectancy at Birth, Female")
```

Life expectancy is greatly influenced by the level of child
mortality. Right after WWII we see the enormous effects of reducing
child mortality. Today, mortality improvement is still steady, but
because the increases in survival are at older ages, the same pace of
improvement in mortality rates adds fewer years of life expectancy at
birth.


Q1.2 What is the pace of life expectancy improvement from 1980 to to
2010?

A. About 1 year per year.

B. About 2-3 years per decade

C. About 1 month per year

```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer1.2 = NA
check(answer1.2)
```


Now we are ready to explore the history of Japan's age-structure.

```{r}
## read in matrix of population counts
tmp <- read.table(file = "http://courses.demog.berkeley.edu/goldstein175/Basket/japan_pop_matrix_1947_2015.txt")
colnames(tmp) <- gsub("X", "", colnames(tmp))
Nxt.mat <- as.matrix(tmp)
# ages 0-110 are rows of the matrix, years 1947-2015 are on the columns
dim(Nxt.mat)
# births for each year are given in the row labeled "0"
print(Nxt.mat["0",]) # or equivalently 'print(Nxt.mat[1,])'

```
Q1.3 What is happening to the number of births from 2014 to 2015?

A. They are increasing as a residual effect of past population growth.

B. They are declining at a rate of 1 or 2 percent a year


```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer1.3 = NA
check(answer1.3)
```


Let's now visualize the change in Japan's age structure. We start by
viewing the change frame-by-frame:

```{r}

year.vec <- colnames(Nxt.mat)
for (i in 1:length(year.vec))
{
    barplot(Nxt.mat[,i], horiz = T,
            xlim = c(0, 2.5 * 10^6),
            xlab = "Number of people N(x,t)", ylab = "Age x")
    title(year.vec[i])
##     Sys.sleep(1/10)
}

```

Or we can compare a few snapshots every 20 years.

```{r}
my.year.vec <- c("1950", "1970", "1990", "2010")
par(mfrow = c(1,4)) # this tells graphics window to display 1 row of 4 panels
for (i in 1:length(my.year.vec))
{
    print(i)
    this.year <- my.year.vec[i]
    barplot(Nxt.mat[,this.year], horiz = T,
            xlim = c(0, 2.5 * 10^6))
    title(my.year.vec[i])
}

```


If we look at the effects of fertility decline on the population
age-structure, we can see a very young population in the 1950s,
followed by a dramatic drop in the fraction of the young in the 1970s
and 1980s. Since the 1980s, the smaller birth cohorts have entered the
workforce, and the ratio of the elderly to workers has risen sharply.

From the point of view of population age-structure, is Japan worse off
today than it was right after WWII? From a well-being point of view,
we care about the social and economic roles of people of different
ages. Demographers use dependency ratios as summary statistics of the
population age-distribution.. We will use the following conventions,
where we define the youth as being aged 0-19, the workers aged 20-64,
and old-age'd as aged 65+.

* Youth Dependency Ratio: $ YDR = N(0-19) / N(20-64)$
* Old-age Dependency Ratio: $ OADR = N(65+) / N(20-64)$
* Total Dependency Ratio: $ TDR = YDR + OADR$

We will use R to loop through each year and create vectors for the
numbers of youth, workers, and elderly by year.

```{r}

x <- 0:110                              # ages

young.vec <- apply(Nxt.mat[x %in% 0:19,], 2, sum)
worker.vec <- apply(Nxt.mat[x %in% 20:64,], 2, sum)
old.vec <- apply(Nxt.mat[x %in% 65:110,], 2, sum)

ydr.vec <- young.vec/worker.vec
oadr.vec <- old.vec/worker.vec
tdr.vec <- oadr.vec + ydr.vec
plot(year.vec, tdr.vec)

```

Q1.4 Wait, is something wrong?

A. Yes, the total dependency ratio can not be greater than 1.0 -- we made
a mistake.

B. Yes, the total dependency ratio even with the contemporary, very aged
population, is lower than it was before fertility decline. We made a
mistake.

C. No. This could be right.



```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer1.4 = NA
check(answer1.4)
```

Let's now look at the three dependency ratios on the same graph:

```{r}

plot(year.vec, oadr.vec, ylim = c(0, 1.5),
     type = "l", col = "red", lwd = 3,
     ylab = "Dependents per worker")
lines(year.vec, tdr.vec, col = "black", lwd = 3)
lines(year.vec, ydr.vec, col = "blue", lwd = 3)
text(1960, .4, "Youth \n Dep Ratio",
     col = "blue")
text(1970, 1, "Total \n Dep Ratio",
     col = "black")
text(1960, .2, "Old \n Dep Ratio",
     col = "red")

```

Now we can see that the youth dependency ratio has fallen more than
the old-age dependency ratio has risen. Total dependency has not yet
reached it's post-war highs.

Q1.5 Why did the youth dependency ratio fall sooner than the old-age
dependency ratio rises?

A. Because when births fall, there is a lag of a few decades between
fewer babies and fewer workers.

B. Because total dependency fell and then rose.

C. Because mortality fell before fertility


```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer1.5 = NA
check(answer1.5)
```



# Part 2. Japan's Future?

We now project Japan's population (using the 1-sex projection model we
covered in lecture) to see how the history of low
fertility will play out over time. We assume here that fertility will
remain at current levels. This will tend to produce quite a bit of
population aging, more so that would occur if, as many expect,
fertility were to rebound somewhat. On the other hand, we also assume
no continued increase in longevity -- which is almost surely
unrealistic -- and thus our projection will have fewer survivors to
old age than is likely.  (Note: we also assume zero net migration --
which we do for simplicity, but which also may not be implausible.)

We begin by importing the projection matrix that made from 2014
rates found in the Human Fertility and Human Mortality Databases.

```{r}
## you can skip the details on how this is read in
tmp <- read.table("http://courses.demog.berkeley.edu/goldstein175/Basket/japan_projection_matrix_2014.txt")
colnames(tmp) <- gsub("X", "", colnames(tmp))
A.2014 <- as.matrix(tmp)
## but do let's look at the first row
print("Fertility rates: row 1 of the projection matrix")
print("(look for non-zero values at ages where women tend give birth)")
print(round(A.2014[1,],3))                       # [1,] indexes first row
```

And let's print some elements showing survivors
```{r}
print(A.2014[1:6, 1:6]) # the upper, left corner
print(A.2014[90:96, 90:96]) # towards the lower, right corner
```

To see the whole matrix is a bit difficult since it is so large. You
can scroll. But you can also do a 3-d picture.

```{r}
persp(x = 0:110,
      y = 0:110,
      z = t(A.2014),                    # transpose for cleaner view
      phi = 35,                         # viewing angle
      xlab = "Age ('in')",
      ylab = "Age ('out')",
      col = "pink",
      expand = .2,
      zlab = "Transition rate")
```

Here, the "fin" along the diagonal is survivorship and the very little
ridge in front is the first row for reproduction.



Q2.1 The sum of the first row elements should be closest to

A. Life expectancy

B. The total fertility rate, the average number of children a woman
would have if she survived through childbearing.

C. The net reproduction rate, the average number of newborn girl would
have would have, accounting for her probability of dying before she
completes childbearing.

D. The average number of daughters the average woman would have if she
survived through childbearing.

```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer2.1 = NA
check(answer2.1)
```


## Projection

Here is some code to project the 2014 population up to 2017.


```{r}
## create a new matrix to store the past and future population in by age
x <- 0:110
year.vec.new <- 1947:2100
Nxt.mat.new <- matrix(NA,
                      nrow = length(x),
                      ncol = length(year.vec.new))
dimnames(Nxt.mat.new) <- list(x, year.vec.new)
Nxt.mat.new[, year.vec.new %in% year.vec] <- Nxt.mat

## check to see how full it is
image(y = x,
      x = year.vec.new,
      z = t(Nxt.mat.new))
## another view (you can play with phi and theta values to change angle)
persp(y = x,
      x = year.vec.new,
      z = t(Nxt.mat.new),
      zlab = "Population N(x,t)",
      xlab = "Time t",
      ylab = "Age x",
      phi = 70, theta = -25)
```

Now we project up to 2017.

```{r}
future.year.vec <- 2015:2017
for (i in 1:length(future.year.vec))
{
    ## The Nxt matrix is indexed by columns with years as characters
    ## so we to refer to N[,"2017"], not N[,2017] (which would be the
    ## 2017th column)
    ## To make this work, we first manipulate the years as numbers and
    ## then convert them to characters before using them as indices.
    to.year = future.year.vec[i]
    from.year = to.year - 1
    to.string <- as.character(to.year)
    from.string <- as.character(from.year)
    ## Now we can do our projection
    print(to.string)
    Nxt.mat.new[,to.string] <- A.2014 %*% Nxt.mat.new[,from.string]
}
```

And we can view the results in several ways.

```{r}
image(y = x,
      x = year.vec.new,
      z = t(Nxt.mat.new))
abline(v = 2014, col = "black", lwd = 1, lty = 2)

persp(y = x,
      x = year.vec.new,
      z = t(Nxt.mat.new),
      phi = 60, theta = -20)
```


### Analysis of Projection (your turn)

For this section, you should create new r-code chunks.

1. Project the population using the same "2014" projection matrix from
   2017 to 2100. (You can do this by copying the code above and making
   some small modifications.)

2. Make a plot of the youth, old-age, and total dependency ratios from
   1947 to 2100. (You can do this by copying the code we used for
   calculating and graphing the intitial population Nxt.mat).


Graded Question 1.

A. Submit a copy of your graph of the dependency ratios from 1947
to 2100. You can put your name on it by using the command

> text(x = 2050, y = 1, "Alfred E. Newman")

B. Will workers will have a harder time supporting dependents in 2100
than in 1947? Please give a yes or no answer. But do state what you
are assuming about the burden of different aged dependents. (Either
answer can be correct, as long as it is consistent with your stated
assumption.) [ < 25 words]



# Part 3. Adding detail with refined age-schedules

Do kids really "cost" the same as the elderly? Does a 20 year old
produce as as much as a 50 year old? Does a 90 year old need more care
than a 70 year old?

Survey data enables age-specific estimates of consumption and
production by age. The methodology for creating these estimates is
somewhat complex, but the findings are quite robust. Here we take
advantage of the schedules that were used by Ron Lee et al. (http://science.sciencemag.org/content/sci/suppl/2014/10/08/346.6206.229.DC1/Lee.SM.pdf) in
order to examine at a finer scale the effects of age structure on the
economy.

Let's look at the estimates of production and consumption by single years
of age in Japan.

```{r}
df <- read.table("http://courses.demog.berkeley.edu/goldstein175/Basket/japan_nta_schedules.txt")
x <- 0:110
cx <- df$cx                             # consumption by age
yx <- df$yx                             # production by age
## (techical note 1: this "yx" is actually labor income).
## (technical note 2: we don't have data over age 90, so we extrapolated)
yx[x > 90] <- 0
cx[x > 90] <- 0

million = 10^6
plot(x, cx,
     type = "l", col = "red",
     ylim = c(0, 5 * million),
     ylab = "cx, yx",
     xlab = "age x",
     main = "Consumption and production, \n Japan c. 2010")
lines(x, yx, type = "l", col = "blue")
```

Note: the decline to zero at age 90 and beyond is because there is
missing data. (We discuss this the effect of this missing data in
Question 5 at the end of the lab, but for now we will assume the true
values are zero.)

Q3.1 What are plausible units for the y-axis

A. Annual dollars per person by  year of age

B. Life-time dollars per person up to that age

C. Annual yen per person by  year of age

D. Life-time yen per person up to that age

```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer3.1 = NA
check(answer3.1)
```


Q3.2 At what age do Japanese first produce as much as they consume?

A. About age 45

B. About age 25

C. About age 20

D. About age 60


```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer3.2 = NA
check(answer3.2)
```

Q3.3 True or False. The schedules imply that the Japanese population
consumes much more than they produce, since the area below the red
line is much larger than the area below the blue line.

A. True.

B. False.

```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer3.3 = NA
check(answer3.3)
```


### Support Ratios

We can apply these more refined age-schedules of economic activity to
the Japanese population that we have already observed and forecast
from 1947 to 2100.

The support ration in year $t% is defined as
$$
SR(t) = { \sum_x N(x,t) y(x,t) \over \sum_x N(x,t) c(x,t)},
$$

where
* $N(x,t)$ is the number of people aged $x$ in year $t$,
* $y(x,t)$ is the production per person aged $x$ in year $t$,
and
* $c(x,t)$ is the consumption per person aged $x$ in year $t$.

In what follows, we assume that $y(x,t)$ and $c(x,t)$ remain constant
at the levels observed in about 2010.

We now calculate the support ratio over the entire period,
1947-2100. (Just make sure that you have done your projection already,
so that your "Nxt.mat.new" is already filled in completely.)

```{r}
SR.vec <- NULL
for (i in 1:length(year.vec.new))
{
    total.production.of.population <- sum(yx * Nxt.mat.new[,i], na.rm = T)
    total.consumption.of.population <- sum(cx * Nxt.mat.new[,i], na.rm = T)
    SR.vec[i] <- total.production.of.population/
        total.consumption.of.population
}
plot(year.vec.new, SR.vec)
```

Q3.4 Taking into account the detailed profiles of consumption and
production by age, what does Support Ratio show about dependency in
2100 compared to 1947?

A. The Support Ratio is lower in 2100 than in 1947, so this means less
dependency.

B. The SR is lower in 2100 than in 1947, so this means there are fewer
producers per equivalent-unit consumer, effectively *increasing*
dependency.

```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer3.4 = NA
check(answer3.4)
```


# Part 4. Optimimum fertility (using stable populations)

In this final section of the lab, we try to reproduce some of the
results on optimal fertility in the Lee et al. reading.

We follow Lee et al in optimising is the "support ratio",
but this time with stable opulation age structures

$$
SR = { \sum_x n_r(x) y(x) \over \sum_x n_r(x) c(x)},
$$

where $n_r(x)$ is proportion of  people aged $x$ in the stable
pouplation with exponential growth rate $r$, $y(x)$ is the
production per person aged $x$ observed in 2010, and $c(x)$ is the
consumption per person aged $x$ observed in 2010.

In a stable population, we derived in class the result that
$$
n_r(x) = b e^{-rx} \ell(x),
$$
where $n_r(x)$ is the proportion of people aged $x$ in a stable
population with exponential growth rate $r$.

Substituting, the stable age formula gives us
$$
SR(r) = { \sum_x e^{-rx} \ell(x) y(x) \over \sum_x e^{-rx} \ell(x) c(x)},
$$

We can now use this formula to study how the support ratio will vary
with the growth rate of the population and with fertility.

Let's begin by specifying a vector with growth rates corresponding to
different total fertility

```{r}
## let TFR range from 1 to 3, in 1/10 of a child increments
tfr.vec <- seq(from = 1, to = 3, by = .1)
## calculate implied pop growth rate "r"
r.vec <- log(tfr.vec / 2.05) / 30
## Here, we divide number of children of both sexes by 2.05 to get the
## number of daughters (since about 105 boys are born per 100
## girls). The generation length is assumed to be 30 years.
plot(tfr.vec, r.vec)
```

Now let's calculate the support ratio $SR(r)$ for each value of
exponential growth.

First we need a survival curve. For this we use the subdiagonal of
the projection matrix, which is a close enough for our purposes.
```{r}
## (no need to understand this)
surv.subdiagonal.of.projection.matrix <- diag(A.2014[-1, -ncol(A.2014)])
Lx <- cumprod(c(1, surv.subdiagonal.of.projection.matrix))
## check to make sure it gives reasonable life expectancy
e0.check <- sum(Lx)
print(e0.check)
```


```{r}
SR.vec <- NULL
for (i in 1:length(r.vec))
{
    r <- r.vec[i]
    nx.r <- exp(-r * x) * Lx
    total.production.of.population <- sum(yx * nx.r)
    total.consumption.of.population <- sum(cx * nx.r)
    SR.vec[i] <- total.production.of.population/
        total.consumption.of.population
}
## let's plot as a function of "r"
plot(r.vec, SR.vec, axes = T)
## and again as a function of tfr
plot(tfr.vec, SR.vec, axes = T)
## now we can add a line at the maximum by eyeballing and guessing the
## value
abline(v = 2.6)
## or more exactly
opt.tfr <- tfr.vec[which.max(SR.vec)]
abline(v = opt.tfr, col = "red")
print(opt.tfr)
```

We see here that the "optimal" fertility rate is greater than 2.


Q4.1 This value corresponds to which value of table 2 on page 232 of in Lee et al.

A. 1.34

B. 2.70

C. 2.33

D. 1.88

E. 1.57

```{r}
## Replace the NA with your answer (e.g., 'A' in quotes)
answer4.1 = NA
check(answer4.1)
```

The other thing to note is the Support Ratio with optimal fertility
rate is not that much higher than that obtained with plus or minus half a
child or so. The curve $SR(r)$ has a very flat peak. This is because
when comparing across stable populations, the increase in the old from
slower pouplation growth is largely offset by decrease in the young.


# Part 5. Another Capital Idea

To finish up, let's combine the fertility on  population age
structure, with the effects of fertility on the population growth rate
and capital accumulation. We can use the Solow model once
again to gain some insights into both the qualititative and the
quantitative effects of population growth rates.

There are possible many ways to do this. (Those interested in the
details can consult the "supplementary materials" of the Lee et al
paper in Science).  For the lab, we will reproduce the "low cost capital
approach," which assumes that the capital-output ratio stays constant
over time. (This is what we already saw Piketty building upon.)

In the Solow model, we saw that if one wants to perserve a constant
stock of capital per person, faster population growth rates need to be
matched with larger fractions of income saved.

In the steady state, we have savings per person equaling the new
capital needed to keep up with population growth $n$, depreciation
$d$, and technological advances $\lambda$,
$$
s y = (n + d + \lambda) k
$$

We can rearrange this steady-state to give us.

$$
s = (n + d + \lambda) (k / y)
$$

Lee et al.  assume standard values such that $k/y \approx 3.0$, $d
= 0.05$, and $\lambda = .02$. This allows calculation of $s$ as a
function only of $n$.

```{r}
d = .05
lambda = .02
ky.ratio = 3.0
s.vec = (r.vec + d + lambda) * ky.ratio
plot(r.vec, s.vec)
```

To take account of the effect of increased savings on the support
ratio, we reduce the effective production of workers by a factor of
$(1-s)$. This operationalizes the idea that the more savings is
required, the less produciton is available for consumption.

$$
SR_{capital} = { \sum_x (1- s_r) e^{-rx} \ell(x) y(x) \over \sum_x e^{-rx} \ell(x) c(x)},
$$

where we use the savings rate $s_r$ that corresponds to the
population growth rate $r$.

```{r}

## This is  the "low capital" case, where the capital-income ratio is
## fixed at 3.0. This is would happen if we had Cobb-Douglas.

## In this case, the a share "s" needs to be set aside from income as
## savings, and cannot be consumed. So we exclude this share from the
## numerator of the support ratio.

SR.low.capital.vec <- NULL

## let's get the savings rate for each growth rate
s.vec <- (r.vec + d + lambda) * ky.ratio
plot(tfr.vec, r.vec)
for (i in 1:length(r.vec))
{
    r <- r.vec[i]
    my.Nx <- Lx * exp(-r * x)
    s <- s.vec[i]
    SR.low.capital.vec[i] <-  sum((1- s) * my.Nx * yx) / sum(my.Nx * cx)
}
plot(tfr.vec, SR.low.capital.vec)
opt.tfr.low.capital <- tfr.vec[which.max(SR.low.capital.vec)]
print(opt.tfr.low.capital)
abline(v = opt.tfr.low.capital, col = "red")
```


# Graded Questions

> 1. See above

> 2. If a population sees a large fertility decline in year "0"

> A. About how many years later would the total dependency ratio reach a
minimum?  [1 sentence is fine]

> B. About how long after this would the TDR stay low? [1 sentence is fine]

> C. Explain why such a low TDR could not be achieved in a stable
>	population simply by varying the exponential growth
>	rate. [ About 2 sentences]

> 3. We compared the effect of a very young age-structure in 1947 with
>    the very old age structure in 2100 using the Total Dependency
>    Ratio and the Support Ratio. The SR calculation suggested that
>    indeed the aged population in 2100 would be a greater economic
>    burden than the young pouplation in 1947. But the TDR calculation
>    did not. What feature of the consumption profile do you think is
>    responsible for this result?  [2 or 3 sentences is fine]

> 4. If robotization were able to make the economy even more capital
>    intensive, say increasing the capital-income ratio from 3.0
>    to 5.0., would the optimal fertility rate be higher or lower than
>    we found above? Explain briefly [< 50 words]

> 5. Our data for Japanese consumption and production ends at age 90
>    (and here we assumed both were zero after this age). What do you
>    think would happen to optimum fertility if we extended both
>    schedules observed at age 90 out to 110? Would optimal fertility be
>    higher or lower? (Hint: look at your graph of the age schedules)
>    [<50 words]

> 6. Does including capital in a model of the economy make pro-baby
>    politics more attractive or less attractive? Explain based on the
>    results of the final part of the lab. [1 sentence is fine]


Congratulations! You are finished with Lab 6.




