---
title: "Filling in some of the details on optimal matching for marriage"
author: Joshua R. Goldstein
output: html_notebook
---

Here I give a bit more detail on the examples in lecture.

1. I show the 6 possible different marriage combinations from our 3x3 example.

2. I show how to use the functional form $Z = h(x^r + y^r)^{1/r}$ to get complements or substitutes in optimal sorting.

# Our 3x3 example

We had the example

```{r}
Z = matrix(byrow = T, nrow = 3, c(5, 8, 2,
                                  7, 9, 6,
                                  2, 3, 0))
colnames(Z) = paste0("F", 1:3)
rownames(Z) = paste0("M", 1:3)
print(Z)
```

Now we can illustrate the different possible marriage arrangements

```{r}
marry.plot <- function(Z, m.vec, f.vec, ...)
{
  ## this is a bit ugly because R puts row 1 of matrix at bottom of the plot, 
  ## whereas we humans like it at the top.
  uZ <- Z[rev(1:nrow(Z)),] # upside down Z
  plot(col(uZ), row(uZ), # pch = as.character(uZ),
       type = "n",
       axes = F, ylab = "", xlab = "")
  lines(f.vec, rev(m.vec), type = "c", ...) # upside down 
  text(col(uZ), row(uZ), round(uZ,2))
  my.match <- uZ[cbind(rev(m.vec), f.vec)] # select matches
  my.sum <- sum(my.match)
  axis(2, at = 1:3, labels = paste0("M",3:1), las = 2,
       tick = FALSE)
  axis(3, at = 1:3, labels = paste0("F", 1:3), tick = FALSE)
  title(sub =paste("Sum =", my.sum), line = 1, ...)
}

par(mfrow = c(2,3) )
marry.plot(Z, 1:3, 1:3)
marry.plot(Z, 1:3, c(1,3,2))
marry.plot(Z, 1:3, c(2,3,1), col.sub = "red", col = "red")
marry.plot(Z, 1:3, c(2,1,3))
marry.plot(Z, 1:3, c(3,1,2))
marry.plot(Z, 1:3, c(3,2,1))

```

# Complements and Substitutes

The following household production function can produce positive or negative assortative marriage, depending on the parameter $r$. We assume $r$, $A_i$, and $A_j$ are all positive.

$$
Z = h(A_i, A_j) = \left( A_i^r + A_j^r \right)^{1 / r}
$$

The cross-derivative is (if I haven't made a mistake):
$$
{\partial^2 Z \over \partial A_i \partial A_j} = (1 - r)(A_i A_j)^{r-1} (A_i^r + A_j^r)^{{1\over r} -2}.
$$
The only term that can change sign is $(1-r)$, with the rest of the expression always being positive, as long as $A_i$ and $A_j$ are positive. So,

$$
{\partial^2 Z \over \partial A_i \partial A_j} > 0, \quad \mbox{if} \quad r < 1
\qquad \rightarrow Complements
$$

and

$$
{\partial^2 Z \over \partial A_i \partial A_j} < 0, \quad \mbox{if} \quad r > 1 
\qquad \rightarrow Substitutes.
$$

## Example 1: Substitutes and Negative Sorting

We can use this production function to produce examples of pay-off matrices $Z$ that lead to positive or negative optimal sorting.

Let's order the men and the women so that there attribute value is equal to 1 for M1 and F1, 2 for M2 and F2, and 3 for M3 and F3. So the pairing M1 and F2 with r = 2 would give
$$
Z_{12} = h(1,2) = (1^2 + 2^2)^{1/2} = \sqrt{5}
$$
We can fill out the whole matrix 
```{r}
h <- function(x,y,r) {(x^r + y^r)^(1/r)}

Z <- matrix(NA, 3,3)
m.vec <- 1:3
f.vec <- 1:3
dimnames(Z) <- list(paste0("M",m.vec), paste0("F",f.vec))
for (i in 1:3)
  for (j in 1:3)
    Z[i,j] = h(i,j, r = 2) # note: r = 2
print(round(Z,1))
```

We can see which one is the maximum using our code from before

```{r}
Z <- round(Z,2)
par(mfrow = c(2,3))
marry.plot(Z, 1:3, 1:3)
marry.plot(Z, 1:3, c(1,3,2))
marry.plot(Z, 1:3, c(2,3,1))
marry.plot(Z, 1:3, c(2,1,3))
marry.plot(Z, 1:3, c(3,1,2))
marry.plot(Z, 1:3, c(3,2,1), col.sub = "red", col = "red")
```
Here we have $r = 2$. According to our calculus result, $r > 1$ means that we should see a negative cross-derivative, the attributes will be substitutes and so "opposites will attract." The maximum match pairs M1 with F3 and M3 with F1, consistent with result.

# Example 2: Complentarity

If we let $r = 1/2 < 1$, we should have complementarity, and so the optimum match should pair likes with likes.

```{r}
Z <- matrix(NA, 3,3)
m.vec <- 1:3
f.vec <- 1:3
dimnames(Z) <- list(paste0("M",m.vec), paste0("F",f.vec))
for (i in 1:3)
  for (j in 1:3)
    Z[i,j] = h(i,j, r = 1/2) # note: r = 1/2
print(round(Z,1))
```

```{r}
Z <- round(Z,2)
par(mfrow = c(2,3))
marry.plot(Z, 1:3, 1:3, col.sub = "red", col = "red")
marry.plot(Z, 1:3, c(1,3,2))
marry.plot(Z, 1:3, c(2,3,1))
marry.plot(Z, 1:3, c(2,1,3))
marry.plot(Z, 1:3, c(3,1,2))
marry.plot(Z, 1:3, c(3,2,1))
```
So for $r < 1$, we see, as we should, that the characteristics are complements.

# A question: what happens if $r = 1$?

If $r = 1$, $Z = A_i + A_j$, so what would we expect to happen?

You can cut and paste the above code (making sure to set r = 1) to check your answer. 


