---
title: "R Notebook for Population Projection: In-class Lab"
output: html_notebook
---

# Overview

In this in-class lab our goal is to project a population with 3 age-groups using matrix multiplication. We are trying to learn several things:

1. How matrix multiplication takes account of births and deaths in an age-structured pop

2. Constant age-specific rates --> a steady state ("stable" population) 

3. "Stable" populations have constant exponential growth rates and age distributions

(Note: at the end of this lab is a section we will use next week when we look at the demographic transition with changing age-specific rates.)

# 0. Population projection as a set of transitions

Let's draw on the board the transitions between 

* young (aged 0-19)
* middle aged (aged 20-39)
* old (aged 40-59)

from the year 2000 to the year 2020.

We should now have a "connected graph", showing us the transitions between states.

Let's put the same information in an "Input-Output" table, with input-states on the columns, and output-states on the rows.

            | Young (in) | Middle (in) | Old (in) |
-----------------------------------------------------
Young (out)  |            |             |          |
Middle (out) |            |             |          |
Old (out)    |            |             |          |


* Which of these table entries could be positive?
* Which will be "structural" zeros?




# 1. Matrix equation for population projection

$$
{\bf A} \cdot {\bf P}(t) = {\bf P}(t +n)
$$

where
* $n$ is projection step length
* $A$ is the projection matrix
* $P$ is a column vector of population by age group

Writing out the matrices,

$$
\begin{pmatrix} F_I & F_{II} & F_{III} \\
S_{I} & 0 & 0 \\ 
0 & S_{II} & 0
\end{pmatrix}
\begin{pmatrix}
P_{1}(t) \\ P_{II}(t) \\ P_{III}(t)
\end{pmatrix}
= 
\begin{pmatrix}
P_{1}(t+n) \\ P_{II}(t+n) \\ P_{III}(t+n)
\end{pmatrix}
$$

Let's multiply out on the board, and compare to our "connected graph" of the transitions.

Let's make sure we understand that our matrix multiplication is doing what we want.

# 2. An example

Let's take a projection matrix with a positive long-run growth rate

$$
{\bf A} = 
\begin{pmatrix} .6 & .9 & 0 \\
.9 & 0 & 0 \\ 
0 & .7 & 0
\end{pmatrix}
$$
How many daughters would be produced by a new-born girl over her lifetime? If the answer is > 1, positive long-term growth.

Let's start with a young'ish population with more young than old.

$$
{\bf P}(0) = 
\begin{pmatrix}
50 \\ 30 \\ 20
\end{pmatrix}
$$
Let's answer the following questions (first by hand, ..., and then we can check with 'R')

A. What is population after one projection step, $P("1")$?

B. What is ratio of total population size at "1" to time "0"? 

C. What is the annual population growth rate from "0" to "1" if we assume a projection step is 20 years?


## A. 1 projection step
```{r}

A = matrix(c(.6, .9, 0,
            .9, 0, 0,
            0, .7, 0), ncol = 3, byrow = T)
P0 = matrix(c(50,
            30,
            20), ncol = 1, byrow = T)

P1 = A %*% P0  # '%*%' means matrix multiplication
print(P1)

```

## B. Ratio of population sizes
```{r}
P0.total = sum(P0)
P1.total = sum(P1)
my.ratio = P1.total / P0.total
print(my.ratio)
```
## C. Annual growth rate
```{r}
R.0.to.1 = log(P1.total/P0.total)/20
print(R.0.to.1)
```

# 3. Longer-term projection

Let's say that demographic rates stay the same, and so the projection matrix stays the same. What happens in the long-term to 
* the rate of population growth?
* the age structure?

We can use the computer to project a longer time, say 10 steps.

```{r}

## approach 1: repeated multiplication

P1 = A %*% P0
P2 = A %*% P1
# etc .....
```
```{r}
## approach 2: Iteration
P.mat = matrix(NA, nrow = 3, ncol = 10+1) # create an empty pop matrix
P.mat[,1] = P0 # initialize by putting starting population in first column
for (i in 1:10) # loop
{
 P.mat[,i+1] = A %*% P.mat[,i] #result goes in i+1 column
}

## let's plot the results
matplot(1:11, t(P.mat), type = "o", main = "Age groups, regular scale")
```
Looks like exponential growth. We can check if we get straight lines in log-scale
```{r}
matplot(1:11, t(P.mat), log = "y", type = "o", main = "Age groups, log scale")
```
## What is the "stable" growth rate?

Let's plot the total population for each time period
```{r}
P.total.vec = NULL
for (i in 1:11)
 P.total.vec[i] = sum(P.mat[,i])
plot(1:11, P.total.vec)
```
Now, let's calculate exponential growth rate and plot
```{r}
R.vec = diff(log(P.total.vec))/ 20
plot(R.vec, type = "b")
```
and print out
```{r}
print(round(R.vec, 3)) ## looks like long term growth rate is about 1.1 percent
```

## What happens to the age stucture?

Let's do an "animated" population pyramid
```{r}
for (i in 1:11)
 barplot(P.mat[,i], main = paste("step", i), names.arg = c("y", "m", "o"), horiz = T)
```

# 4. Conclusions:

1. Population projection with a constant set of demographic rates leads to constant exponential growth

2. The age-structure settles down to a steady state, which we call the "stable age structure"

We now close R-studio and 

* Look at the stable age structures implied by real demographic rates (in an App)

http://shiny.demog.berkeley.edu/josh/stable/

* Review everything using continuous time and an analytical approach


# 5. Final section (for next week)

Population momentum and the Demographic Transition 


If we have a growing population, and put the breaks on fertility, so that fertility is at "replacement", what will happen to population size?

Let's say we have a new matrix ${\bf A_0}$, where every woman has a total of 1 daughter over her lifetime. The long-term growth rate will be 0, right?
$$
{\bf A_0} = 
\begin{pmatrix} 0.5 & 0.5& 0 \\
1 & 0 & 0 \\ 
0 & 1 & 0
\end{pmatrix}

$$

We can check the long-term growth rate by guessing the stable age-structure as having equal numbers at each age.

```{r}
A0 = matrix(c(.5, .5, 0,
             1, 0, 0,
             0, 1, 0), byrow = T, ncol = 3)
P0.guess = matrix(c(10,10,10), byrow = T, ncol = 1)
P.mat = matrix(NA, nrow = 3, ncol = 10+1) # an empty pop matrix
P.mat[,1] = P0.guess # the starting population
for (i in 1:10)
{
 P.mat[,i+1] = A0 %*% P.mat[,i]
}

## let's plot the results
matplot(1:11, t(P.mat), type = "o", main = "Age groups, regular scale")
```

Now let's simulate decline in fertility by starting with the age-structure of our growing population, but applying it to the zero-growth demographic rates?
```{r}

P.mat = matrix(NA, nrow = 3, ncol = 10+1) # an empty pop matrix
P.mat[,1] = P0 # our old "young'ish" starting population
for (i in 1:10)
{
 P.mat[,i+1] = A0 %*% P.mat[,i] # we use A0 (not A)
}
## let's get population total
P.total.vec = NULL
for (i in 1:11)
{
 P.total.vec[i] = sum(P.mat[,i])
}

## let's plot the results
matplot(1:11, P.total.vec, type = "o", main = "Total population, population momentum")
```

What is happening here?

We have "transitional dynamics", drop in fertility rates reduces number of young at first, but then the age-pyramid fills up, giving us some continued population growth

Let's look at an animation:

Let's do an "animated" population pyramid
```{r}
for (i in 1:11)
 barplot(P.mat[,i], main = paste("step", i), names.arg = c("y", "m", "o"), horiz = T, 
         xlim = c(0, 100))
```

We see that the total number of young has gone down a bit, but that age-structure has changed.

When would the age-structure be best from an economic point of view? (Hint: look at first 3 time steps)

