# The Great Recession and Fertility in the United States


To begin with let's look at the plot of historical fertility rates in
the United States.

Q1. Does a better economy ...

A. Always lead to higher fertility

B. Sometimes lead to higher fertility and sometimes lead to lower
fertility

C. Usually lead to higher fertility

D. I see no relationship?


Q2. In terms of our cost-of-time model, if a better economy raised
fertility, then this would mean

A. The substitution effect dominates the income effect

B. The income effect dominates the substitution effect


Q3. If men and women become more alike, then what would we expect?

A. Stronger substitution effects

B. Stronger income effects

C. It's not so clear ...

Q4. What do we expect in the Great Recession?

A. Lower fertility

B. Higher fertility

C. No change

Q5. What do we expect now that the Great Recession is over?

A. Fertility to return to it's previous level?

B. Fertility to be even higher, making up for lost time.

C. Fertility to stay lower, until people are really sure the recovery
will continue.

# Data analysis

We will look at unemployment rates and fertility rates.

## Importing the data

```{r}
prefix <- "http://courses/demog.berkeley.edu/goldstein175/Basket"
usa.df <- read.csv(url = paste0(prefix,"usa_tfr_ue_by_year.csv"))
state.df <- read.csv(url = paste0(prefix,"state_tfr_ue_by_year.csv"))
```

## The Great Recession

### Nationally

```{r}

## save ourselves some typing
tfr <- usa.df$tfr
ue <- usa.df$ue
year <- usa.df$year                     # 2001:2015
## to compare UE in year of conception
ue.before <- ue[year %in% 2001:2014]
tfr.after <- tfr[year %in% 2002:2015]
```

We can plot the relationships
```{r}

par(mfrow = c(2,2))
plot(year, tfr, type = "o",
     col = "blue",
     main = "Total Fertility Rates, USA")
plot(year, ue, type = "o",
     col = "red",
     main = "Unemployment, USA")
plot(ue, tfr, type = "p",
     col = "black",
     main = "TFR by unemployment, USA")
abline(lm(tfr ~ ue))
plot(ue.before, tfr.after,
     main = "TFR by prior year's unemployment")
abline(lm(tfr.after ~ ue.before))
```

Q. Is higher unemployment associated with lower fertility?


A. Yes, throughout the whole period

B. No, higher unemployment is associated with higher fertility.

C. It depends on which period we look at

D. I have no idea!

Q. Does it make a difference if we look at conditions at the time of
conception vs. time of birth?

A. Not much


Let's look at the period up to 2011, separately.

```{r}

s <- year %in% 2001:2011
par(mfrow = c(2,2))
plot(year, tfr, type = "o",
     col = "blue",
     main = "Total Fertility Rates, USA")
points(year[s], tfr[s], pch = 19, col = "grey")
plot(year, ue, type = "o",
     col = "red",
     main = "Unemployment, USA")
points(year[s], ue[s], pch = 19, col = "grey")
plot(ue, tfr, type = "p",
     col = "black",
     main = "TFR by unemployment, USA")
points(ue[s], tfr[s], pch = 19, col = "grey")
## abline(lm(tfr ~ ue))
abline(lm(tfr.after ~ ue.before, subset = s),
       col = "grey", lty = 2)
plot(ue.before, tfr.after,
     main = "TFR by prior year's unemployment")
## abline(lm(tfr.after ~ ue.before))
abline(lm(tfr.after ~ ue.before, subset = s),
              col = "grey", lty = 2)
points(ue.before[s], tfr.after[s], pch = 19, col = "grey")
```

So, we see that the recession seems to have lowered fertility on a
national level.

We can also look at the states to see if states that were harder hit,
saw a bigger decline in fertility.

Import a fancy function to plot an panel of states in the shape of the USA
```{r}
## old: source("recession_functions_for_class.R")
source(paste0(prefix,"recession_functions_for_class.R"))
```

Now display the state TFR and UE.

```{r}

state.panel.plot(x = state.df$year,
                 y = state.df$tfr,
                 state = state.df$state.name,
                 vline = 2008,
                 hline = 2.0,
                 new = T,
                 type = "l",
                 col = "black",
                 legend.ylab = "tfr",
                 legend.xlab = "year",
                 title = "TFR by state, 2001-2015, all",
                 ylim = c(1.5, 3))
state.panel.plot(x = state.df$year,
                 y = state.df$tfr.filt,
                 state = state.df$state.name,
                 vline = 2008,
                 hline = 2.0,
                 col = "blue",
                 type = "l",
                 lwd = 2,
                 new = F,
                 ylim = c(1.5, 3))

```

Now look at UE

```{r}
## now do unemployment
state.panel.plot(x = state.df$year,
                 y = state.df$ue,
                 state = state.df$state.name,
                 vline = 2008,
                 hline = c(.05, .1),
                 col = "red",
                 type = "l",
                 lwd = 2,
                 new = T,
                 ylim = c(0, .15))

```


```{r}
## now do a scatter plot
par(c(1,1))
plot(state.df$ue,
     state.df$tfr.filt, type = "p")
m0 <- lm(tfr.filt ~ ue , data = state.df)
abline(m0)
```

Q. Do "fracking states" see less effect of Recession?

Q. Does their fertility also not fall as much?

Q. How about coastal states?

Q. Which states have biggest increase in unemployment?

Q. Which states have biggest decline in fertility?

Overall, it seems like income effect dominates. 200 years after
Malthus, are we still in his regime?

## Concluding discussion question:

Going back to national plot:

Why hasn't TFR recovered as UE has gone down?

How could we test your theories?



